---
title: "CHST_midterm"
author: "Lisa Chemplayil, Nicholas Huron, Nicolina Sciarretta, Tom Thomas"
date: "3/1/2018"
output: html_document
---

```{r setup, results='hide', message=FALSE}
#Authors: TT, NGS, NAH

#clear workspace first
rm(list=ls())

knitr::opts_chunk$set(echo = TRUE)
#set the directory
knitr::opts_knit$set(root.dir = "/Volumes/GoogleDrive/My Drive")
#For PCs
#knitr::opts_knit$set(root.dir = "G:/My Drive/")

#load required packages
library(tidyverse)
library(ggmap)
library(RColorBrewer)
library(sp)
library(raster)
library(rgdal)
library(maptools)

#set working directory as needed if the root.dir call fails
#For mac
#setwd("/Volumes/GoogleDrive/My Drive/")
#For PCs
#setwd("G:/My Drive/")
```

##I. WRANGLE the imap data
###Read in invasives dataset and polygon shapefiles

We load all of the shapefiles for now, just in case we need any of them later.
```{r read data, results='hide', message=FALSE}
#Authors: TT, NGS, NAH

#read in invasives
invasives <- read_csv("./QuantSci_GP/data/PA_Invasive_Species/all_obs_imap_18Dec17_v2_0.csv", progress = FALSE)

#read in shapefiles
town <-readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles", layer='PaMunicipalities2017_01')
s_senate  <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles", layer='PaSenatorial2017_01')
s_house   <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles", layer='PaHouse2017_01')
congress  <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles", layer='PaCongressional2017_01')
county    <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles", layer='PaCounty2017_01')
```

###Clean up the read in data (Part 1)

We rename the lat/long coordinate columns for ease of reference.
```{r rename latlong}
#Authors: TT

colnames(invasives)[colnames(invasives) %in% 'obsorigxcoord'] <- 'long'
colnames(invasives)[colnames(invasives) %in% 'obsorigycoord'] <- 'lat'
```

Now, we need to make sure that all of the coordinate systems play nicely by standardizing the coordinate reference systems for shapefiles.
```{r standardize shapefile CRS}
#Authors: TT

county <- spTransform(county, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
s_senate <- spTransform(s_senate, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
congress <- spTransform(congress, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
town <- spTransform(town, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
s_house <- spTransform(s_house, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
```

###Identify and note data contained in each SpatialPolygonsDataFrame. Select, arrange, rename, or mutate data as needed.

We focus on the grain of federal congressional districts. In the SpatialPolygonsDataFrame object `congress`, which contains information on the 18 districts. The `data` section contains the following columns that are relevant for our analyses:
  1. LEG_DISTRI [factor]: the unique number identifier for each district
  5. PARTY  [factor]: the one letter code for the political party of each district's representative
  
This object also contains a `polygons` section that gives spatial data that helps draw each district. The `ID` identifies each polygon to a district (although it starts counting at zero and connects through the `plotOrder` section). There is also a `bbox` section that defines the plot area and a `proj4string` section that contains the coordinate reference system to translate latitude and longitude values to the plot.


###Filter out data on the spotted lanternfly from `invasives`

First, we should confirm the best way to filter out, either by scientific name or common name. To do this, we will summarise the data that have common names containing the word `"fly"`, just to make sure nothing gets lost in syntax variation.
```{r EDA lantern fly}
#Authors: NAH

invasives %>%
  group_by(state_scientific_name, stateCommonName) %>%
  count() %>%
  filter(str_detect(stateCommonName, pattern = "fly")) %>%
  arrange(desc(n))
```

>It is clear that there are a few observations with fly, and it looks like `"Spotted Lanternfly"` is the consistent common name for this species, but we will stick to the scientific name to be certain when we `filter` it out below in our new `invasives` object (called `invasives2`).

```{r filter}
#Authors: TT, NAH

#get original row count
nrow(invasives)

invasives2 <- invasives %>%
  filter(state_scientific_name != "Lycorma delicatula")

#check that we lost some rows
nrow(invasives2)
```

###Clean up the read in data (Part 2)

We know from previous communications that the `invasives$obsdate` variable contains data that are not equally reliable. To be certain that we use dates with reliable years only, we are going to clean `obsdate` and then `mutate()` a new variable (called `year`) that we will use for temporal analyses. To do so, we can search the `obscomments_long` entries for rows from `obsdate` that need tidying!

```{r clean obsdate}
#Authors: NAH

#explore obsdate and obscomments_long
invasives2 %>%
  dplyr::select(obsdate, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(d|D)ate")) %>%
  arrange(obscomments_long)

#get number of rows of interest
invasives2 %>%
  dplyr::select(obsdate, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(d|D)ate")) %>%
  arrange(obscomments_long) %>%
  nrow()
```

We now can confirm that there are 724 rows that contain some reference to the date in the comments section. These likely represent dates that need fixing or may need to be removed from the dataset. We also realize we need to format the dates at some point. First, let us find the true unknowns/uncertains by using the keywords `"unknown"` and `"placeholder"`/`"place-holder"`.
```{r find unknowns}
#Authors: NAH

invasives2 %>%
  dplyr::select(obsid, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(d|D)ate")) %>%
  filter(str_detect(obscomments_long, pattern = "(u|U)nknown | (p|P)lace[\\-holder]*")) %>%
  arrange(obscomments_long)
```

Of the 724 rows, only 56 have the keywords we are worried about. These 56 rows represent dates that may be slated for removal. The other 668 are likely ambiguous but can be determined to year with confidence. Now we have to trim the former before verifying the years for `obsdate` values in the latter. We can use `obsid` values as our identifiers for the rows we need to remove (manually spot-checked the 56 rows).
```{r rm questionable date rows}
#Authors: NAH

#check the nrows of invasives2 to start
nrow(invasives2)

#object of rows to remove with certainty
rows2rm <- c("PA-6898U", "PA-379U", "PA-8018U", "PA-6643U", "PA-6407U", "PA-6885U",	"PA-6766U", "PA-6891U", "PA-6693U",	"PA-6761U")

#remove these rows from invasives2
invasives2 <- invasives2 %>%
  dplyr::filter(!(obsid %in% rows2rm))

#check that rows were dropped
nrow(invasives2)
```

As it turns out, only 10 of these rows lack a confident year of observation, so we removed them from this dataset. The rest may need year spot checking... but for now, we  can create a new column, `year` that contains the year of each obsdate.

```{r mutate years}
#Authors: NAH

invasives2 %>%
  dplyr::mutate(year = str_extract(obsdate, "\\d{4}"))
```


###Verifying Dates and clean up the read in data (part 3)
Using `View()` of `invasivescheck` below, one can check that dates are correct in the table of just the specified columns. With that table you can search for keywords in columns. In the comments column the word "date" was first searched, with over 700 entries. If one picked repetivitive words from thse 700 entries that . The next word search was "representative", which resulted in 110 entries. Of those 110 entries about half are dated in `obsdate` and `year` as 1980, however, in comments its states that "1980 is a repesentative of the time frame (1980-2007) when this county was known to begin hosting HWA". The next word search was "unknown" these resulted in 31 entries, 13 of these refer to the exact day/month is unkown but year is known. The last search word was "known", this also brought up "unknown" again. "known" had 218 entries, abut 50 of the entries refered to the "representative" word, and more than half refer to unknown day/month but known year.
```{r check years}
#Authors: NGS

View(invasivescheck <- invasives2 %>% dplyr::mutate(year = str_extract(obsdate, "\\d{4}")) %>% 
  dplyr::select(obsid, stateCommonName, obsdate, obscomments_long, year))

```

According to these observations the years do match the comments, however we need to remove the observations that have a date range of 1980-2007. 
```{r remove representative}
#Authors: NAH

#filter out the observations with the 1980-2007 date range. This can be done with a match of "representative", but when observing the filtration of View(), there are some other unintended rows that are fine for dates. Instead, let's filter with that date range in the comments, "1980-2007" as well!
(rows2rm2 <- invasives2 %>%
  dplyr::select(obsid, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(r|R)epresentative")) %>%
  filter(str_detect(obscomments_long, pattern = "1980\\-2007")) %>%
  arrange(obscomments_long) %>%
  dplyr::select(obsid))

rows2rm2 <- as_vector(rows2rm2)

#invasives2 row check
nrow(invasives2)

#remove these rows from invasives2
invasives2 <- invasives2 %>%
  dplyr::filter(!(obsid %in% rows2rm2))

#final row check in invasives2
nrow(invasives2)

```

##II. EDA - Univariate  
###Identify focal species

First, we read in the `iMap_spp.csv` file to a new object `imap_spp`. We can trim the new object down to all entries we already tidied by selectively joining `imap_spp` to `invasives2` by `"state_scientific_name", "stateCommonName", and "natlhabitat"`. The resulting object is `invasives3`. 
```{r read in imap_spp and join}
#Authors: NAH

#read in imap_spp
imap_spp <- read_csv("./QuantSci_GP/data/PA_Invasive_Species/iMap_sppp.csv")
#change the colname to match invasives2
colnames(imap_spp)[colnames(imap_spp) %in% "SpeciesType"] <- "speciestype"

#left join to invasives2 to make new object, invasives3
invasives3 <- left_join(x = invasives2, y = imap_spp, by = c("state_scientific_name", "stateCommonName", "natlhabitat"))

#now we need to drop the invasives2 version of speciestype and rename speciestype.y to speciestype
invasives3 <- invasives3 %>%
  dplyr::select(-speciestype.x, speciestype = speciestype.y)

invasives3

```

Now, data exploration can be done to select focal taxa. Full disclosure, an exploration of this summary was done previously, leading to the use of aquatic species, as they are taxonomically diverse and less biased towards overwhelming plant observations. The final filtered version of just **aquatic invasive species** is called `aqua_invasives`.
```{r look at aquatic v terrestrial and then filter}
#Authors: NAH
imap_spp %>%
  #filter(state_scientific_name %in% as_vector((invasives3 %>% distinct(state_scientific_name)))) %>%
  filter(state_scientific_name %in% unique(invasives3$state_scientific_name)) %>%
  group_by(natlhabitat) %>%
  count()
imap_spp %>%
  filter(state_scientific_name %in% unique(invasives3$state_scientific_name)) %>%
  group_by(natlhabitat, speciestype) %>%
  count() %>%
  arrange(desc(n))

#filter for natlhabitat == "aquatic"
(aqua_invasives <- invasives3 %>%
  filter(natlhabitat == "aquatic" | natlhabitat == "Aquatic"))

```

>It is easy to see that working with aquatic species may involve fewer species, but the involved species are separated by group more evenly (not 201 of 209 as plants, as in terrestrial species).


We also need to make the `aqua_invasives` object a SpatialPolygonsDataFrame and set the CRS for that as well, now that we have tidied it. To make this easier, we will create a new object, `aqua_invasives_spatial`. This way, we can always reference the original tibble.
```{r convert invasives}
#Authors: TT, NAH
aqua_invasives_spatial <- aqua_invasives

coordinates(aqua_invasives_spatial) <- ~long+lat
proj4string(aqua_invasives_spatial) <- projection('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')
```
  


###Species Richeness Per Federal Congressional
```{r join SR and congressional data}
#Authors: NGS
srPerCongressional <- over(aqua_invasives_spatial, congress)
srPerCongressional <- cbind(srPerCongressional, aqua_invasives_spatial$state_scientific_name)
colnames(srPerCongressional)[colnames(srPerCongressional) %in% 'aqua_invasives_spatial$state_scientific_name'] <- 'species' 
srPerCongressional <- group_by(srPerCongressional, LEG_DISTRI) 
srPerCongressional <- dplyr::summarise(srPerCongressional, sr=length(unique(species)))
```

```{r plot the results, message = FALSE} 
#Authors: NGS
invasivesCongressional <- congress
invasivesCongressional@data <- left_join(invasivesCongressional@data, srPerCongressional, by='LEG_DISTRI') 


invasivesCongressionalgg <- fortify(invasivesCongressional, region='LEG_DISTRI') 
invasivesCongressionalgg <- left_join(invasivesCongressionalgg, invasivesCongressional@data, by=c("id" = 'LEG_DISTRI')) 

p <- ggplot(data=invasivesCongressionalgg)+
  geom_polygon(aes(x=long, y=lat, group=group, fill=sr))+ 
  coord_fixed(1.3)+ 
  scale_fill_gradient() +
  ggtitle(str_to_title("Species Richness of Invasive Aquatic Species In Pennsylvania")) +
    labs(y = "Latitude of PA in Degrees", x = "Longitute of PA in Degrees")

p
```

Now, we can break down aquatic invasive species richness by congression LEG_DISTRI for the five major species types in the dataset:
```{r split by species type}
#Authors: NGS, NAH
turtles <- aqua_invasives %>% filter(speciestype == "turtle")
plants <- aqua_invasives %>% filter(speciestype == "plant")
arthropods <- aqua_invasives %>% filter(speciestype == "arthropod")
mollusc <- aqua_invasives %>% filter(speciestype == "mollusc")
fish <- aqua_invasives %>% filter(speciestype == "fish")

#combine into a list for easy looping
byspeciestype <- list(turtles, plants, arthropods, mollusc, fish)
names(byspeciestype) <- c("turtles", "plants", "arthropods", "mollusc", "fish")
```

Now that we have filtered versions of `aqua_invasives` for each `speciestype`, so we need to turn them into **SpatialPolygonsDataFrames** and calculate **Species Richness (SR)**. In summarizing the **SR** for each `speciestype`, we need to populate counties that are missing in the output with **SR** set to zero. This is done mid-loop, check the code comments for further details.

```{r loop get SR by LEG_DISTRI, message = FALSE}
#Authors: NGS, NAH
#create new version of list that contains each species type
spatial_byspeciestype <- byspeciestype

#loop to set the coordinate system and reference system
for(a in seq_along(spatial_byspeciestype)){
  coordinates(spatial_byspeciestype[[a]]) <- ~long+lat
  proj4string(spatial_byspeciestype[[a]]) <- projection('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')
}

#next, we need to perform a similar loop to overlay the spatial invasives data with the congressional LEG_DISTRI data. we will also end up summarising SR by LEG_DISTRI for each group in this loop.

#initialize holder lists for loops  
sr_per_congressional_byspeciestype <- spatial_byspeciestype
invasive_congressional_byspeciestype <- spatial_byspeciestype
invasive_congressional_gg_byspeciestype <- spatial_byspeciestype
  
for(b in seq_along(spatial_byspeciestype)){
  sr_per_congressional_byspeciestype[[b]] <- over(spatial_byspeciestype[[b]], congress)
  sr_per_congressional_byspeciestype[[b]] <- cbind(sr_per_congressional_byspeciestype[[b]], spatial_byspeciestype[[b]]$state_scientific_name)
  colnames(sr_per_congressional_byspeciestype[[b]])[colnames(sr_per_congressional_byspeciestype[[b]]) %in% 'spatial_byspeciestype[[b]]$state_scientific_name'] <- 'species'
  sr_per_congressional_byspeciestype[[b]] <- group_by(sr_per_congressional_byspeciestype[[b]], LEG_DISTRI)
  sr_per_congressional_byspeciestype[[b]] <- dplyr::summarise(sr_per_congressional_byspeciestype[[b]], sr=length(unique(species))) %>% 
    filter(!is.na(LEG_DISTRI))
  
#we need to add code here that adds the missing districts with SR==0
if(length(setdiff(congress$LEG_DISTRI, sr_per_congressional_byspeciestype[[b]]$LEG_DISTRI)) > 0){
sr_per_congressional_byspeciestype[[b]] <- add_row(sr_per_congressional_byspeciestype[[b]], LEG_DISTRI = setdiff(congress$LEG_DISTRI, sr_per_congressional_byspeciestype[[b]]$LEG_DISTRI), sr = rep(0, length(setdiff(congress$LEG_DISTRI, sr_per_congressional_byspeciestype[[b]]$LEG_DISTRI))))
}
  
  invasive_congressional_byspeciestype[[b]] <- congress
  invasive_congressional_byspeciestype[[b]]@data <- left_join(invasive_congressional_byspeciestype[[b]]@data, sr_per_congressional_byspeciestype[[b]], by = "LEG_DISTRI")
  
  invasive_congressional_gg_byspeciestype[[b]] <- fortify(invasive_congressional_byspeciestype[[b]], sr_per_congressional_byspeciestype[[b]], region = 'LEG_DISTRI')
  invasive_congressional_gg_byspeciestype[[b]] <- left_join(invasive_congressional_gg_byspeciestype[[b]], invasive_congressional_byspeciestype[[b]]@data, by = c("id" = 'LEG_DISTRI'))
}

```

Now we have the object `invasive_congressional_gg_byspeciestype` that is formatted to plot the map of results for each species

```{r loop to plot by group} 
#Authors: NGS, NAH
#loop through ggplot commands
for(c in seq_along(invasive_congressional_gg_byspeciestype)){
   print(
    ggplot(data = invasive_congressional_gg_byspeciestype[[c]]) +
    geom_polygon(aes(x=long, y=lat, group=group, fill=sr))+ 
    coord_fixed(1.3)+ 
    scale_fill_gradient()+
    ggtitle(str_to_title(paste0("Species Richness of Invasive Aquatic ", names(invasive_congressional_gg_byspeciestype)[c], " in Pennsylvania"))) +
    labs(y = "Latitude of PA in Degrees", x = "Longitute of PA in Degrees")
   )
}

```


###Univariate Distributions of richness for each grain and group

To compare species richness between grains, we need to develop a version of the code for `congress` and `aqua_invasives` for all other grain types. Before we do that, let's show we can plot the univariate curve of SR for congress

```{r plot sr curve congress}
#Authors: NAH
ggplot(data = srPerCongressional) +
  geom_density(aes(sr), fill = "black") +
  ggtitle(str_to_title(paste0("Species Richness of Invasive Aquatic Species"))) +
  labs(y = "Density", x = "sr")
```

Let's create a list with the SR for all grains:
```{r get SR per grain}
#Authors: NAH

#town    MUNICPAL_
#s_senate LEG_DISTRI
#s_house LEG_DISTRI
#congress LEG_DISTRI
#county  COUNTY_NAM

#initialize a bunch of lists to hold  results
sr_per_grain <- list(NULL, NULL, NULL, NULL, NULL)
grains <- list(town, s_senate, s_house, congress, county)
names(grains) <- c("town", "s_senate", "s_house", "congress", "county")
names(sr_per_grain) <- rep(paste0("srPer_", names(grains)))
names(sr_per_grain) <- names(grains)


#we can loop for the three grains with LEG_DISTRI (We will need to manual entry for the other two after that)
for(d in seq_along(sr_per_grain)[2:4]){
  sr_per_grain[[d]] <- over(aqua_invasives_spatial, grains[[d]])
  sr_per_grain[[d]] <- cbind(sr_per_grain[[d]], aqua_invasives_spatial$state_scientific_name)
  colnames(sr_per_grain[[d]])[colnames(sr_per_grain[[d]]) %in% 'aqua_invasives_spatial$state_scientific_name'] <- 'species'
  sr_per_grain[[d]] <- group_by(sr_per_grain[[d]], LEG_DISTRI)
  sr_per_grain[[d]] <- dplyr::summarise(sr_per_grain[[d]], sr=length(unique(species)))
}

#manual for town
  sr_per_grain[[1]] <- over(aqua_invasives_spatial, grains[[1]])
  sr_per_grain[[1]] <- cbind(sr_per_grain[[1]], aqua_invasives_spatial$state_scientific_name)
  colnames(sr_per_grain[[1]])[colnames(sr_per_grain[[1]]) %in% 'aqua_invasives_spatial$state_scientific_name'] <- 'species'
  sr_per_grain[[1]] <- group_by(sr_per_grain[[1]], MUNICIPAL_)
  sr_per_grain[[1]] <- dplyr::summarise(sr_per_grain[[1]], sr=length(unique(species)))

#manual for county
  sr_per_grain[[5]] <- over(aqua_invasives_spatial, grains[[5]])
  sr_per_grain[[5]] <- cbind(sr_per_grain[[5]], aqua_invasives_spatial$state_scientific_name)
  colnames(sr_per_grain[[5]])[colnames(sr_per_grain[[5]]) %in% 'aqua_invasives_spatial$state_scientific_name'] <- 'species'
  sr_per_grain[[5]] <- group_by(sr_per_grain[[5]], COUNTY_NUM)
  sr_per_grain[[5]] <- dplyr::summarise(sr_per_grain[[5]], sr=length(unique(species)))
  
```

Now we can plot density plots to compare grains:
```{r plot sr curve congress}
#Authors: NAH

#plot the curves together
#note that we could benefit from changing the alphas in the legend
ggplot() +
  geom_density(data = sr_per_grain[[1]], aes(sr, fill = names(sr_per_grain)[1]), alpha = 0.5) +
  geom_density(data = sr_per_grain[[2]], aes(sr, fill = names(sr_per_grain)[2]), alpha = 0.5) +
  geom_density(data = sr_per_grain[[3]], aes(sr, fill = names(sr_per_grain)[3]), alpha = 0.5) +
  geom_density(data = sr_per_grain[[4]], aes(sr, fill = names(sr_per_grain)[4]), alpha = 0.5) +
  geom_density(data = sr_per_grain[[5]], aes(sr, fill = names(sr_per_grain)[5]), alpha = 0.5) +
  ggtitle(str_to_title(paste0("Species Richness Of Invasive Aquatic Species By Grain"))) +
  labs(y = "Density", x = "sr") +
  scale_fill_manual(name = "Grain", values = c("red", "blue", "green", "orange", "black"), labels = c("congress", "county", "s_house", "s_senate", "town"), guide = 'legend')
```