---
title: "NAH_QSGP_midterm_notebook"
author: "Nicholas Huron"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
#Authors: TT, NGS, NAH

#clear workspace first
rm(list=ls())

knitr::opts_chunk$set(echo = TRUE)
#set the directory
knitr::opts_knit$set(root.dir = "/Volumes/GoogleDrive/My Drive/")
#For PCs
#knitr::opts_knit$set(root.dir = "G:/My Drive/")

#load required packages
library(tidyverse)
library(ggmap)
library(RColorBrewer)
library(sp)
library(raster)
library(rgdal)
library(maptools)
library(lubridate)

#set working directory as needed
#For mac
#setwd("/Volumes/GoogleDrive/My Drive/")
#For PCs
#setwd("G:/My Drive/")
```

##I. WRANGLE the imap data
###Read in invasives dataset and polygon shapefiles

We load all of the shapefiles for now, just in case we need any of them later.
```{r read data}
#Authors: TT, NGS, NAH

#read in invasives
invasives <- read_csv("./QuantSci_GP/data/PA_Invasive_Species/all_obs_imap_18Dec17_v2_0.csv")

#read in shapefiles
town <-readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles/", layer='PaMunicipalities2017_01')
s_senate  <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles/", layer='PaSenatorial2017_01')
s_house   <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles/", layer='PaHouse2017_01')
congress  <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles/", layer='PaCongressional2017_01')
county    <- readOGR(dsn = "./QuantSci_GP/data/PA_Invasive_Species/shapefiles/", layer='PaCounty2017_01')
```

###Clean up the read in data (Part 1)

We rename the lat/long coordinate columns for ease of reference.
```{r rename latlong}
#Authors: TT

colnames(invasives)[colnames(invasives) %in% 'obsorigxcoord'] <- 'long'
colnames(invasives)[colnames(invasives) %in% 'obsorigycoord'] <- 'lat'
```

Now, we need to make sure that all of the coordinate systems play nicely by standardizing the coordinate reference systems for shapefiles.
```{r standardize shapefile CRS}
#Authors: TT

county <- spTransform(county, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
s_senate <- spTransform(s_senate, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
congress <- spTransform(congress, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
town <- spTransform(town, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
s_house <- spTransform(s_house, CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'))
```

###Filter out data on the spotted lanternfly

First, we should confirm the best way to filter out, either by scientific name or common name. To do this, we will summarise the data that have common names containing the word `"fly"`, just to make sure nothing gets lost in syntax variation.
```{r EDA lantern fly}
#Authors: NAH

invasives %>%
  group_by(state_scientific_name, stateCommonName) %>%
  count() %>%
  filter(str_detect(stateCommonName, pattern = "fly")) %>%
  arrange(desc(n))
```

>It is clear that there are a few observations with fly, and it looks like `"Spotted Lanternfly"` is the consistent common name for this species, but we will stick to the scientific name to be certain when we `filter` it out below in our new `invasives` object (called `invasives2`).

```{r filter}
#Authors: TT, NAH

#get original row count
nrow(invasives)

invasives2 <- invasives %>%
  filter(state_scientific_name != "Lycorma delicatula")

#check that we lost some rows
nrow(invasives2)
```

###Clean up the read in data (Part 2)

We know from previous communications that the `invasives$obsdate` variable contains data that are not equally reliable. To be certain that we use dates with reliable years only, we are going to clean `obsdate` and then `mutate()` a new variable (called `year`) that we will use for temporal analyses. To do so, we can search the `obscomments_long` entries for rows from `obsdate` that need tidying!

```{r clean obsdate}
#Authors: NAH

#explore obsdate and obscomments_long
invasives2 %>%
  dplyr::select(obsdate, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(d|D)ate")) %>%
  arrange(obscomments_long)

#get number of rows of interest
invasives2 %>%
  dplyr::select(obsdate, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(d|D)ate")) %>%
  arrange(obscomments_long) %>%
  nrow()
```

We now can confirm that there are 724 rows that contain some reference to the date in the comments section. These likely represent dates that need fixing or may need to be removed from the dataset. We also realize we need to format the dates at some point. First, let us find the true unknowns/uncertains by using the keywords `"unknown"` and `"placeholder"`/`"place-holder"`.
```{r find unknowns}
#Authors: NAH

invasives2 %>%
  dplyr::select(obsid, obscomments_long) %>%
  filter(str_detect(obscomments_long, pattern = "(d|D)ate")) %>%
  filter(str_detect(obscomments_long, pattern = "(u|U)nknown | (p|P)lace[\\-holder]*")) %>%
  arrange(obscomments_long)
```

Of the 724 rows, only 56 have the keywords we are worried about. These 56 rows represent dates that may be slated for removal. The other 668 are likely ambiguous but can be determined to year with confidence. Now we have to trim the former before verifying the years for `obsdate` values in the latter. We can use `obsid` values as our identifiers for the rows we need to remove (manually spot-checked the 56 rows).
```{r rm questionable date rows}
#Authors: NAH

#check the nrows of invasives2 to start
nrow(invasives2)

#object of rows to remove with certainty
rows2rm <- c("PA-6898U", "PA-379U", "PA-8018U", "PA-6643U", "PA-6407U", "PA-6885U",	"PA-6766U", "PA-6891U", "PA-6693U",	"PA-6761U")

#remove these rows from invasives2
invasives2 <- invasives2 %>%
  dplyr::filter(!(obsid %in% rows2rm))

#check that rows were dropped
nrow(invasives2)
```

As it turns out, only 10 of these rows lack a confident year of observation, so we removed them from this dataset. The rest may need year spot checking... but for now, we  can create a new column, `year` that contains the year of each obsdate.

```{r mutate years}
#Authors: NAH

invasives2 %>%
  dplyr::mutate(year = str_extract(obsdate, "\\d{4}"))
```

We also need to make the `invasives2` object a SpatialPolygonsDataFrame and set the CRS for that as well, now that we have tidied it.
```{r convert invasives}
#Authors: NAH

coordinates(invasives2) <- ~long+lat
proj4string(invasives2) <- projection('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')
```


  2. Identify and note data contained in each SpatialPolygonsDataFrame. Select, arrange, rename, or mutate data as needed before working through tasks II and III.
