---
title: "Week2B_Huron"
author: "Nicholas A. Huron"
date: 2018-02-06
output: html_document
---

#Chapter 5

```{r setup, include=F}
knitr::opts_chunk$set(echo = T)
#required packages
library(nycflights13)
library(tidyverse)
```

##Section 5.2
###1. Filter the flights dataset according to the following conditionals:

1.1: arrival delay >= 120 min
```{r 5.2.1.1}
filter(flights, arr_delay >= 120) #10200
```

1.2: flew to Houston (IAH or HOU)
```{r 5.2.1.2}
filter(flights, dest %in% c("IAH", "HOU")) #9313
```

1.3: operated by united, american, or delta
```{r 5.2.1.3}
#find the codes
unique(flights$carrier)
#filter
filter(flights, carrier %in% c("AA","UA", "DL")) #139504
```

1.4: departed in summer (July, August, September)
```{r 5.2.1.4}
filter(flights, month > 6 & month < 10) #86326
```

1.5: arrive > 120 min late but did not depart late
```{r 5.2.1.5}
filter(flights, arr_delay > 120 & dep_delay <= 0) #29
```
1.6: delayed >= 60 min and made up over 30 min in flight (difference in initial and final delays >30)
```{r 5.2.1.6}
filter(flights, dep_delay >= 60 & (dep_delay - arr_delay) > 30) #1844
```

1.7: departed between 00:00 and 06:00 inclusive ( dep_time >= 0000 and dep_time <=0600)

```{r 5.2.1.7}
filter(flights, dep_time >= 0 & dep_time <= 600)  #9344
```

##Section 5.3
###4. Which flights travelled the longest? Which travelled the shortest?

Flights that travel the longest
```{r 5.3.4a}
arrange(flights, desc(distance))
```

Flights that travel the shortest
```{r 5.3.4b}
arrange(flights, distance)
```

##Section 5.4
###2. What happens if you include the name of a variable multiple times in a select() call?
```{r 5.4.2}
select(flights, dep_time, dep_time, dep_delay, dep_time)
```
When repeated, it does not appear that duplicate columns are printed (regardless of order). Rather, a single column per column name is included.

##Section 5.5
###1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r 5.5.1}
#convert dep_time and sched_dep_time from numeric representation of time to min since midnight
transmute(flights,
          dep_time,
          minutes = ((dep_time %/% 100)*60)+(dep_time %% 100), 
          sched_dep_time, 
          sched_minutes =  ((sched_dep_time %/% 100)*60)+(sched_dep_time %% 100)
          )
```
Now the times are appropriately included here as minutes (even if it makes telling time more difficult)...

##Section 5.6
###5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

First, we group the data by carriers and then we can find the mean departure and arrival delays.
```{r}
#which carrier has the worst delays?
(carriers <- flights %>%
  group_by(carrier) %>%
  summarise(mean_dep_delay = mean(dep_delay, na.rm = T), mean_arr_delay = mean(arr_delay, na.rm = T)))
```

Then we can plot the results
```{r}
#plot
ggplot(data = carriers, mapping = aes(x = carrier)) +
  geom_bar(mapping = aes(y= mean_dep_delay), stat = "identity")
ggplot(data = carriers, mapping = aes(x = carrier)) +
  geom_bar(mapping = aes(y= mean_arr_delay), stat = "identity")
```

Based on these two plots, it sure seems like F9 has the worst delays on average for both arrival and departure.

###Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

1. First, group by carrier and destinations. Summarise the mean values for the dual grouping, removing NA's.
2. Then, re-group by carrier and summarise the standard deviation among destinations and the grand mean.
```{r}
(delays <- flights %>% 
  group_by(carrier, dest) %>% 
  summarise(mean_dep_delay = mean(dep_delay, na.rm = T)) %>% 
  group_by(carrier) %>% 
  summarise(SD = sd(mean_dep_delay, na.rm = T), M = mean(mean_dep_delay, na.rm = T)))
```
We can see the means and SD by dest here, so we can kind of get the picture... Note that the SD is zero for carriers with only one destination.

While it may be possible to potentially get at the dest biases, there is great disparity of flights to airports. Some airlines only go to 1 or 2 destinations, such as F9.

```{r}
#this can be deconstructed
(carriers_dest <- flights %>%
  group_by(carrier, dest) %>%
  filter(!is.na(dep_delay) & dep_delay > 0))
#look at dest and carriers
ggplot(data = carriers_dest) +
  geom_bar(mapping = aes(x = carrier, fill = dest), position = "dodge", show.legend = F)
```

Visually, we can begin to see some of the carriers that perform poorly, regardless of destination quirks (UA comes to mind as a good example).

##Section 5.7
###4. For each destination, compute the total minutes of delay. For each, flight, compute the proportion of the total delay for its destination.

Compute the total minutes of delay.
```{r}
(flights %>%
  group_by(dest) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(total_delay = sum(dep_delay))) 
```
ATL has a ton of delays... I can confirm that this is true after spending a night sleeping there after ~3 weeks of fieldwork.

Compute the proportion of the total delay for each flight's destination.
```{r}
(flights %>%
  group_by(dest) %>%
  filter(!is.na(dep_delay)) %>%
   #get a sum for total delay
  mutate(total_delay = sum(dep_delay)) %>%
   #group by flight numbers
    group_by(flight) %>%
   #now get the proportion for dep_delay
    mutate(prop_delay = dep_delay/total_delay))
```
Prop_delay is our new variable that shows the proportion of a particular flight's delay relative to the total delay for its destination. Note that these values shown at the top of the table are quite small!
