---
title: "Week7B_Ch20t21_Huron.Rmd"
author: "Nicholas Huron"
output: html_document
---

```{r setup,cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
#required packages
library(tidyverse)
```


##Chapter 20

###Exercises 20.3.5
  2. Read the source code for dplyr::near() (Hint: to see the source code, drop the ()). How does it work?
```{r near source code}
dplyr::near
```

The function `dplyr::near()` evaluates the absolute value of the difference between the two input values and assesses whether or not it is less than some tolerance argument, which by default is set to a value related to the floating point of doubles. The output is logical based on assessing this relationship!

###Exercises 20.4.6
  4. Create functions that take a vector as input and returns:

  - The last value. Should you use [ or [[?
```{r lastvalue fn}
#function
lastval <- function(x){
  x[[length(x)]]
}
#test it
zzz <- 1:10
(zzz <- purrr::set_names(x = zzz,nm = letters[1:10]))
lastval(zzz)
```
    
  It seems better to use `[[`, as we only want the value, not the element.

  - The elements at even numbered positions.
```{r even pos only}
evenposonly <- function(x){
  x[seq(from = 2, to = length(x), by = 2)]
}

evenposonly(zzz)
```

  - Every element except the last value.
```{r all but last}
notlast <- function(x){
  x[-length(x)]
}

notlast(zzz)
```

  - Only even numbers (and no missing values).
```{r evensonly}
evensonly <- function(x){
  x[(!is.na(x)) & (x %% 2 == 0)]
}

#throw in an NA to show it working
(zzz <- c(zzz, not = NA))
#and test it
evensonly(zzz)
```
  
###Exercises 20.5.4
  2.

##Chapter 21

###Exercises 21.3.5
3
###Exercises 21.4.1
1
###Exercises 21.5.3
1