---
title: "Week7B_Ch20t21_Huron.Rmd"
author: "Nicholas Huron"
output: html_document
---

```{r setup,cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
#required packages
library(tidyverse)
```


##Chapter 20

###Exercises 20.3.5
  2. Read the source code for dplyr::near() (Hint: to see the source code, drop the ()). How does it work?
```{r near source code}
dplyr::near
```

The function `dplyr::near()` evaluates the absolute value of the difference between the two input values and assesses whether or not it is less than some tolerance argument, which by default is set to a value related to the floating point of doubles. The output is logical based on assessing this relationship!

###Exercises 20.4.6
  4. Create functions that take a vector as input and returns:

  - The last value. Should you use [ or [[?
```{r lastvalue fn}
#function
lastval <- function(x){
  x[[length(x)]]
}
#test it
zzz <- 1:10
(zzz <- purrr::set_names(x = zzz,nm = letters[1:10]))
lastval(zzz)
```
    
  It seems better to use `[[`, as we only want the value, not the element.

  - The elements at even numbered positions.
```{r even pos only}
evenposonly <- function(x){
  x[seq(from = 2, to = length(x), by = 2)]
}

evenposonly(zzz)
```

  - Every element except the last value.
```{r all but last}
notlast <- function(x){
  x[-length(x)]
}

notlast(zzz)
```

  - Only even numbers (and no missing values).
```{r evensonly}
evensonly <- function(x){
  x[(!is.na(x)) & (x %% 2 == 0)]
}

#throw in an NA to show it working
(zzz <- c(zzz, not = NA))
#and test it
evensonly(zzz)
```
  
###Exercises 20.5.4
  2. What happens if you subset a tibble as if you’re subsetting a list? What are the key differences between a list and a tibble?
  
```{r subset a tibble like a list}
as_tibble(iris)[[1]]
as_tibble(iris)[1]
as_tibble(iris)[["Sepal.Length"]]
```

It appears that tibbles respond somewhat similarly to Hadley's description of `[` and `[[` for lists. For a tibble, `[` will call the formatted row as a sub-tibble, much like for a list, it will maintain the list structure with a sub-list. However, when you use `[[`, it returns just the values for both. I imagine that the regular rules of conduct for lists and tibbles apply (e.g., tibbles are properly formatted tables where all cols and rows are of equal length, respectively), whereas lists are free to vary widely in sublist lengths.

##Chapter 21

###Exercises 21.3.5
  3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, show_mean(iris) would print:

    show_mean(iris)
    #> Sepal.Length: 5.84
    #> Sepal.Width:  3.06
    #> Petal.Length: 3.76
    #> Petal.Width:  1.20

(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

So the code below will work, but it isn't as clean as Hadley's...
```{r numeric col means fn}
show_means <- function(x, decimal_digits = 2){
  for(i in seq_along(names(x))){
    if(is.numeric(x[[i]])){ #is_numeric did not work... rather than complicate the conditional, just used is.numeric instead
      hold_mean <- format(x = round(mean(x[[i]]), digits = decimal_digits), nsmall = decimal_digits)
      print(paste0(names(x)[i], ": ", hold_mean))
    } else{
      #do not put anything here, but if we wanted to say so, we could do something like this:
      warning(paste0(names(x)[i]), " does not contain numeric data!")
    }
  }
}

show_means(iris)
```

I guess we can try to do it Hadley's way too?
```{r clean version}
show_means <- function(x, decimal_digits = 2){
#initialize list to store all of the strings
results_show_means <- vector(mode = "list", length = length(names(x)))
  for(i in seq_along(names(x))){
    if(is.numeric(x[[i]])){ #is_numeric did not work... rather than complicate the conditional, just used is.numeric instead
      hold_mean <- format(x = round(mean(x[[i]]), digits = decimal_digits), nsmall = decimal_digits)
      results_show_means[[i]] <- paste0(names(x)[i], ": ", hold_mean)
    } else{
      #do not put anything here, but if we wanted to say so, we could do something like this:
      warning(paste0(names(x)[i]), " does not contain numeric data!\n")
    }
  }
print(results_show_means)
}

show_means(iris)
```
  
###Exercises 21.4.1
  1. 
  
###Exercises 21.5.3
  1. 