---
title: "Week5A_map1_Huron"
author: "Nicholas A. Huron"
date: '2018-02-13'
output:
  pdf_document: default
  html_document: default
---

##Set Up Code with Chunks from Tutorial
###Load Required Packages
This includes setting the working directory with `knitr::opts_knit$set(root.dir=)`.

```{r setup, include=TRUE, results='hide', message=FALSE}
rm(list=ls()) #start with a clean slate!
#required packages from tutorial
require(ggmap); require(sp); require(rgdal); require(raster); require(tidyverse); require(RColorBrewer)
#add viridis package
require(viridis)
knitr::opts_chunk$set(echo = TRUE)
#set the figure dims universally
knitr::opts_chunk$set(fig.width=6, fig.height=10) 
#set the dir
knitr::opts_knit$set(root.dir="/Users/nicholashuron/Google Drive/QuantSci_GP/")
```

###Load Provided Data

1. Load invasive data:
```{r load invasives}
invasives <- read_csv("./data/PA_Invasive_Species/all_obs_imap_18Dec17_v2_0.csv")
```
2. Load raster data
```{r load temp}
temp <- raster("./data/PA_Invasive_Species/raster/paTemp.tif")
```
3. Load polygon shapefile (can also use congressional maps, senate maps, etc. See shapefile folder)
```{r load geopoli polygons}
county <- readOGR(dsn="./data/PA_Invasive_Species/shapefiles/", layer='PaCounty2017_01')
state <- readOGR(dsn="./data/PA_Invasive_Species/shapefiles/", layer='PaState2017_01')
```

###Clean Up the Loaded Data

Change some column names to be more intuitive.
```{r col change}
colnames(invasives)[colnames(invasives) %in% 'obsorigxcoord'] <- 'long'
colnames(invasives)[colnames(invasives) %in% 'obsorigycoord'] <- 'lat'
```

Standardize the projection across all data. Use the temperature data projection (WGS84) as the basis for others
```{r set all to WGS84}
#set the coords in the invasives objects
coordinates(invasives) <- ~long+lat
#set the projection
proj4string(invasives) <- projection(temp)
county <- spTransform(county, CRS(projection(temp)))
state <- spTransform(state, CRS(projection(temp)))
```

###Explore the Data

Let's figure out *the number of locations with invasive species per county!*
```{r invasive locs}
#obtain spatial "counties" values for invasives to confirm counties
invasivesPerCounty <- over(invasives, county)
#summarise by county
invasivesPerCounty <- table(invasivesPerCounty$COUNTY_NAM)
#tibble-fy it
invasivesPerCounty <- as.tibble(invasivesPerCounty)
#rename cols
colnames(invasivesPerCounty) <- c('county', 'nInvasives')
invasivesPerCounty
```

Now how about figuring out the *invasive species richness per county!*
```{r invsive sr}
#same step as above
srPerCounty <- over(invasives, county)
#add the invasive species sci names to the rows
srPerCounty <- cbind(srPerCounty, invasives$state_scientific_name)
#set a new col name to make easier to interpret
colnames(srPerCounty)[colnames(srPerCounty) %in% 'invasives$state_scientific_name'] <- 'species'
#group by county
srPerCounty <- group_by(srPerCounty, COUNTY_NAM)
#and summarise by unique values of species for county
srPerCounty <- dplyr::summarise(srPerCounty, sr=length(unique(species))) # Count number of unique species per county
srPerCounty
```

###Assignment Questions!

####1. Make a barchart of the number of invasive species (species richness) per PA county. Bonus: Can you make the plot in descending order?

We can use the object `srPerCounty` above to make a plot:

```{r sr by county}
ggplot(data = srPerCounty) +
  geom_bar(mapping = aes(x = COUNTY_NAM, y = sr, fill = factor(..x..)), stat = "identity", show.legend=FALSE) +
  coord_flip() +
  labs(y = "Species Richness", x = "County", title = "Species Richness of Invasive Species in PA by County") +
  theme(axis.text.y=element_text(size=8))
```

We can also make in alphabetical order by county (we have to remove NA's to get this to work for now).
```{r sr by county alpha}
ggplot(srPerCounty) +
  geom_bar(mapping = aes(x = COUNTY_NAM, y = sr, fill = factor(..x..)), stat = "identity", show.legend=FALSE) +
  coord_flip() +
  labs(y = "Species Richness", x = "County", title = "Species Richness of Invasive Species in PA by County") +
  scale_x_discrete(limits = rev(levels(srPerCounty$COUNTY_NAM)))
```

Finally, we can keep the NA and do it in descending order by SR for the **BONUS** question!
```{r sr by descending sr}
ggplot(data = srPerCounty) +
  geom_bar(mapping = aes(x = reorder(COUNTY_NAM, sr), y = sr, fill = factor(..x..)), stat = "identity", show.legend=FALSE) +
  coord_flip() +
  labs(y = "Species Richness", x = "County", title = "Species Richness of Invasive Species in PA by County") +
  scale_fill_viridis(discrete = T)
```

\

####2. Make a map that gives the number of invasive species (species richness) per U.S. House of representative district. Hint: You do not have to start from scratch, just cut and paste (and clean) from the tutorial.

```{r fig.height = 6, fig.width = 8}
#so read in the US house district data
house <- readOGR(dsn="./data/PA_Invasive_Species/shapefiles/", layer='PaCongressional2017_01')
#fix it's CRS
house <- spTransform(house, CRS(projection(temp)))

#summarise the SR by district
srPerHouse <- over(invasives, house)
#add the invasive species sci names to the rows
srPerHouse <- cbind(srPerHouse, invasives$state_scientific_name)
#set a new col name to make easier to interpret
colnames(srPerHouse)[colnames(srPerHouse) %in% 'invasives$state_scientific_name'] <- 'species'
#group by county
srPerHouse <- group_by(srPerHouse, LEG_DISTRI) 
#and summarise by unique values of species for county
srPerHouse <- dplyr::summarise(srPerHouse, sr=length(unique(species)))
srPerHouse$LEG_DISTRI <- factor(x = srPerHouse$LEG_DISTRI, levels = sort(as.numeric(srPerHouse$LEG_DISTRI)))
(srPerHouse <- srPerHouse %>%
  filter(!is.na(LEG_DISTRI)) %>%
  arrange(LEG_DISTRI))

# First join the data we put together above with house
invasivesHouse <- house
#make sure levels order matches for the same col that we will join by
invasivesHouse$LEG_DISTRI <- factor(x = invasivesHouse$LEG_DISTRI, levels = sort(as.numeric(invasivesHouse$LEG_DISTRI)))
#now join them by col LEG_DISTRI!
invasivesHouse@data <- left_join(invasivesHouse@data, srPerHouse, by='LEG_DISTRI')
# Map with ggplot
invasivesHousegg <- fortify(invasivesHouse, region='LEG_DISTRI') 
#join the districts and the dataframe id back together
invasivesHousegg <- left_join(invasivesHousegg, invasivesHouse@data, by=c("id" = 'LEG_DISTRI'))

#plot that purtty map
ggplot(data=invasivesHousegg)+
  geom_polygon(aes(x=long, y=lat, group=group, fill=sr)) + # Color the counties by nInvasives
  coord_fixed(1.3) + # This is to fix the height-width scale
  scale_fill_viridis() #i use my own palettes, man
```

####3. Use ggplot (not spplot) to plot the number of invasive occurrences by county by decade from 1900 to the present. Here I would expect about 10-12 plots. Hint: Just complete the code in the tutorial.

Fix the dates, just like in the tutorial:
```{r fix date format}
# First, bins number of invasives over time (bin by decade)
fixDates <- unlist(strsplit(as.character(invasives$obsdate), ' '))
fixDates <- fixDates[seq(1, length(fixDates), by=2)]
invasives$obsdate <- as.Date(fixDates, format='%m/%d/%Y')
invasives$year <- as.numeric(substr(invasives$obsdate, start=1, stop=4))
```

Now we need to subset the data by decade and create new objects by year (invasivesYYYYs) and a summary of observations by county for each decade (invasivesPerCountyYYYYs). We remove `NA`'s first though! See comments for additional steps in the loop.
```{r subset and smush, warning=F}
# Subset the data by decade for bins 
# Note: This code could be made shorter with a 'for loop' or different type of function, 
# Because I am a masochist, let's try to figure out the loop version of this
invasivesNArm <- invasives[! is.na(invasives@data$year),] # Get rid of entries without a year
#set an object for decade start counter for loop
(startdec <- seq(from = 1900, to = 2017, by = 10))
#the loop that gets the invasivesYYYYs objects
for(a in 1:length(startdec)){
  print(paste0("Decade starting with ", startdec[a], " and ending with ", (startdec[a]+9)))
  assign(x = paste0("invasives",startdec[a],"s"), value = invasivesNArm[invasivesNArm@data$year >= startdec[a] & invasivesNArm@data$year <= (startdec[a] + 9),])
#the loop that gets the invasivesPerCountyYYYYs objects and names their cols
  #do the over command to get the county spatial data in
assign(x = paste0("invasivesPerCounty",startdec[a],"s"), value = over(get(paste0("invasives",startdec[a],"s")), county))
  #summarise the data by county name
assign(paste0("invasivesPerCounty",startdec[a],"s"), table(get(paste0("invasivesPerCounty",startdec[a],"s"))$COUNTY_NAM))
  #make it a dataframe
assign(paste0("invasivesPerCounty",startdec[a],"s"), as.data.frame(get(paste0("invasivesPerCounty",startdec[a],"s"))))
  #code that uses a temporary object to rename the columns
temporary <- get(paste0("invasivesPerCounty",startdec[a],"s"))
colnames(temporary) <- c('COUNTY_NAM', 'nInvasives')
#reassign the new col name object back to it's original object
assign(x = paste0("invasivesPerCounty",startdec[a],"s"), value = temporary)
}
```

Now we need to create a version of `county` (see above) for each decade to use. We then create a version for each decade (invasivesCountyYYYYs). Then, the loop stuff craps the bed on us when we try to assign a new value with `left_join()` with the `invasivesCountyYYYYs@data` element of the broader object. So we manually run the `left_join()` command for each decade object, much to my chargrin.
```{r smush}
#now smush together
for(c in 1:length(startdec)){
  assign(x = paste0("invasivesCounty",startdec[c],"s"), value = county)
  #assign(x = paste0("invasivesCounty",startdec[c],"s","@data"), value = left_join(get(paste0("invasivesCounty",startdec[c],"s"))@data, get(paste0("invasivesPerCounty",startdec[b],"s")), by = 'COUNTY_NAM'))  #this line is the problem child... i cannot get assign to assign a value to just the part of invasivesCountyYYYYs@data
}

#manually do the left_join() here
invasivesCounty1900s@data <- left_join(invasivesCounty1900s@data, invasivesPerCounty1900s, by='COUNTY_NAM')
invasivesCounty1910s@data <- left_join(invasivesCounty1910s@data, invasivesPerCounty1910s, by='COUNTY_NAM')
invasivesCounty1920s@data <- left_join(invasivesCounty1920s@data, invasivesPerCounty1920s, by='COUNTY_NAM')
invasivesCounty1930s@data <- left_join(invasivesCounty1930s@data, invasivesPerCounty1930s, by='COUNTY_NAM')
invasivesCounty1940s@data <- left_join(invasivesCounty1940s@data, invasivesPerCounty1940s, by='COUNTY_NAM')
invasivesCounty1950s@data <- left_join(invasivesCounty1950s@data, invasivesPerCounty1950s, by='COUNTY_NAM')
invasivesCounty1960s@data <- left_join(invasivesCounty1960s@data, invasivesPerCounty1960s, by='COUNTY_NAM')
invasivesCounty1970s@data <- left_join(invasivesCounty1970s@data, invasivesPerCounty1970s, by='COUNTY_NAM')
invasivesCounty1980s@data <- left_join(invasivesCounty1980s@data, invasivesPerCounty1980s, by='COUNTY_NAM')
invasivesCounty1990s@data <- left_join(invasivesCounty1990s@data, invasivesPerCounty1990s, by='COUNTY_NAM')
invasivesCounty2000s@data <- left_join(invasivesCounty2000s@data, invasivesPerCounty2000s, by='COUNTY_NAM')
invasivesCounty2010s@data <- left_join(invasivesCounty2010s@data, invasivesPerCounty2010s, by='COUNTY_NAM')
```

Now that we have that done, we can ensure that our spatial data and summary data are joined. Lastly, we use another `for()` loop to plot iteratively for each decade!
```{r some more smushing and then plotting, fig.height = 6, fig.width = 8, warning=F}
for(d in 1:length(startdec)){
  assign(x = paste0("invasivesCountygg",startdec[d],"s"), value = fortify(get(paste0("invasivesCounty",startdec[d],"s")), region='COUNTY_NAM'))
  assign(x = paste0("invasivesCountygg",startdec[d],"s"), value = left_join(get(paste0("invasivesCountygg",startdec[d],"s")), get(paste0("invasivesCounty",startdec[d],"s"))@data, by=c("id"='COUNTY_NAM')))
  #then do the ggplot thang
  print(ggplot(data = get(paste0("invasivesCountygg",startdec[d],"s"))) +
    geom_polygon(aes(x=long, y=lat, group=group, fill=nInvasives)) +
    coord_fixed(1.3) +
    scale_fill_viridis() +
    ggtitle(paste0("Invasive Species Observations by PA County for the ", startdec[d], "s")))
}

```