---
title: "Week10B_Ch25_Huron.Rmd"
author: "Nicholas Huron"
date: "4/3/2018"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
#clear the enviroment
rm(list=ls())
#required packages
library(modelr)
library(tidyverse)
library(gapminder)
library(broom)
#set options
options(na.action = na.warn)
```

##Chapter 25

###Section 25.2.5

  1. A linear trend seems to be slightly too simple for the overall trend. Can you do better with a quadratic polynomial? How can you interpret the coefficients of the quadratic? (Hint you might want to transform year so that it has mean zero.)
  
```{r try a quad poly model}
#test that we can create a quadpoly model and write a function to iterate through
gapminder %>%
  mutate(year = year - mean(year)) %>%
  filter(country == "Afghanistan") %>%
  lm(lifeExp ~ poly(year, 2), data = .)

#make the fnc
country_quadmodel <- function(df) {
  lm(lifeExp ~ poly(year, 2), data = df)
}
```

```{r prepare the dataset}
#transform year so that it has a mean of zero (find mean and subtract it from all values)
#then group and nest
gapminder_quad <- gapminder %>%
  mutate(year = year - mean(year)) %>%
  group_by(country, continent) %>%
  nest()
```

```{r fit the model with map function and fit residuals}

gapminder_quad <- gapminder_quad %>%
  mutate(model = map(data, country_quadmodel)) %>%
  mutate(resids = map2(data, model, add_residuals))

gapminder_quad_resid <- unnest(gapminder_quad, resids)
```

```{r plot results}
gapminder_quad_resid %>%
  ggplot(mapping = aes(x = year, y = resid)) +
  geom_line(aes(group = country), alpha = 1/3) +
  geom_smooth(se = F)

gapminder_quad_resid %>%
  ggplot(mapping = aes(x = year, y = resid, group = country)) +
  geom_line(alpha = 1/3) +
  facet_wrap(~continent)

```

```{r run through the lm model to make comparisons}
#function to iterate
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

#create nested object
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()

#iterate through model
by_country <- by_country %>% 
  mutate(model = map(data, country_model))

#and add residuals
by_country <- by_country %>% 
  mutate(
    resids = map2(data, model, add_residuals)
  )

#unnest with residuals
resids <- unnest(by_country, resids)

```

```{r glance through for model comparisons}
#quad model
quad <- gapminder_quad %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance, .drop = T) %>%
  select(country:r.squared) %>%
  mutate(quad.r.squared = r.squared) %>%
  select(-r.squared)

#linear model
lin <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE) %>%
  select(country:r.squared)

p <- inner_join(quad, lin, by = c("country", "continent")) %>%
  group_by(country) %>%
  arrange(continent) %>%
  ggplot(mapping = aes(x  = r.squared)) +
  geom_density(fill = "blue", alpha = 0.5)
p + geom_density(mapping = aes(x = quad.r.squared), fill = "green", alpha = 0.5)
```

It does appear that the quadratic polynomial model improves the fit of the model to the data (green density curve) relative to the linear one (blue density curve). The coeffcients of the quadratic model can be seen as constants to multiple `year` by to obtain `lifeExp`. The quadratic model will assume certain curvature structure that does not exist in the linear model version.

###Section 25.4.5

  3. What’s missing in the following data frame? How does quantile() return that missing piece? Why isn’t that helpful here?
  
```{r example chunk}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg))) %>% 
  unnest()
```

The quantile probabilities are not included, which causes confusion when unnested! If the above code is amended to the code below, it is easier to associate quantiles and values for each `cyl` with appropriate probabilities.
```{r revised chunk}
probs <- c(0, 0.25, 0.5, 0.75, 1)

mtcars %>% 
  group_by(cyl) %>% 
  summarise(p = list(probs), q = list(quantile(mpg, probs))) %>% 
  unnest()
```

###Part II: What are the differences in the relationship between lifeExp ~ year among continents in the gapminder dataset?

  1. Fit lm models of lifeExp ~ year for each country as described in the ch. 25 text.
  
```{r fit lm models}
#function to iterate
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

#create nested object
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()

#iterate through model
by_country <- by_country %>% 
  mutate(model = map(data, country_model))

confint_tidy(by_country$model[[1]])

```

  2. Use the broom package to make a tidy object of the coefficients (i.e.,the variable named estimate) and their confidence intervals (i.e., the variables named conf.low and conf.high).

```{r get broomtidying}
(
country_mods <- by_country %>%
  mutate(summod = map(model, .f = function(x) tidy(x,conf.int = T))) %>%
  unnest(summod) %>%
  group_by(country, continent) %>%
  arrange(continent)
)
```

  3. Using your own code and style, plot these coefficients along with their confidence intervals for each country grouped by continent using ggplot and one of the facet_* functions such as facet_wrap. 
  
```{r plot}
country_mods %>%
  ggplot(aes(x = country, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank()
        ) +
  facet_grid(term~continent, scales = "free_y")
```

  Bonus: Plot the mean coefficient values for each continent. Plot the confidence intervals around these continent means. (under construction)

```{r test, eval=FALSE}
country_mods %>%
  ungroup() %>%
  group_by(continent, term) %>%
  nest() %>%
  summarise(mean=mean(data$estimate), ci = list(confint(data$estimate)))
```
